---
title: "Conjugate Priors"
author: "Erena Tanabe, Tom Elliott, Matt Edwards"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conjugate Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
h2 {
  border-bottom: 2.5px solid #4682B4;
  padding-bottom: 5px;
}
```


Likelihood functions: inz_lbinom, inz_lmulti, inz_lnorm

Prior functions: inz_dbeta, inz_ddir, inz_dNIG

Posterior calculation: calculate_posterior

Output: summary

To demonstrate how the functions work together, we will use the **SURFIncomeSurvey_200** example dataset from iNZight.

```{r setup}
library(iNZightBayes)
surf_data <- iNZightTools::smart_read("surf.csv")
head(surf_data)
```

<br>

# Beta-Binomial

This is the parameter estimation method for when the primary variable of interest is a two-level categorical variable. Supports single variable and two-variable (categorical/categorical) cases.

With this application we will use the 'inz_lbinom' and 'inz_dbeta' functions to construct our prior and likelihood object.


## Univariate application

### Gender variable

```{r surfGender}
head(surf_data$Gender, n=10)
```

First, we construct our **likelihood** object from the data. This is the likelihood component of the Bayes' Theorem, $p(x | \theta)$.

```{r surfLikelihood}
likelihood <- inz_lbinom(surf_data, Gender)
```

*With this variable, we would like to estimate the proportion of females ('successes') in the population, $\theta$. Hence, we will also be inherently estimating the proportion of males ('failures'), 1-$\theta$, as well.*

Next, we set up **the prior**, by passing the likelihood object into the prior function. This is the prior component of the Bayes' Theorem $p(\theta)$. To start off, we will use an uninformative prior, Beta(1,1). This is the default prior set in the 'inz_dbeta' function.

```{r surfPrior}
prior <- inz_dbeta(likelihood)
```





```{r surfPlot, echo=FALSE}
theta <- seq(0, 1, by=0.001)
y <- dbeta(theta, prior$alpha, prior$beta)

plot(theta, y, type="n", ylim=c(0,1.2),
     main=paste0("Beta(", prior$alpha, ",", prior$beta, ") prior"),
     xlab=expression(theta), ylab="Density", yaxs="i")
polygon(x=c(0,theta,1),
        y=c(0,y,0),
        col="powderblue")
```

*Beta(1,1) = Uniform(0,1). With this prior, we are essentially implying that we don't have prior knowledge of $\theta$ (the proportion of females) in the population. Hence, all possible values of $\theta$ are assigned equal probability.*

Finally, we pass the prior object which contains the likelihood and prior information into the calculate_posterior function. This function calculates the **posterior** parameters. This is the posterior component of the Bayes' Theorem, $p(\theta | x)$.

```{r surfPosterior}
posterior <- calculate_posterior(prior)
```

This posterior object is used to obtain posterior estimates and credible interval values through summary function.

```{r surfSummary}
summary(posterior)
```

As we used an uninformative prior, the estimate of $\theta$ (and 1-$\theta$) is very similar to the proportions seen in the data itself, which is 0.5350 (107/200) for females and 0.4650 (93/200) for males.


### Changing the prior...

We have prior knowledge/belief that $\theta$, the proportion of females in the population is around 0.5 (50%).

Let's change the prior to an informative prior of **Beta(50,50)**:

```{r surfPrior2}
prior <- inz_dbeta(likelihood, alpha = 50, beta = 50)
```

```{r surfPlot2, echo=FALSE}
theta <- seq(0, 1, by=0.001)
y <- dbeta(theta, prior$alpha, prior$beta)

plot(theta, y, type="n",
     main=paste0("Beta(", prior$alpha, ",", prior$beta, ") prior"),
     xlab=expression(theta), ylab="Density", yaxs="i")
polygon(x=c(0,theta,1),
        y=c(0,y,0),
        col="powderblue")
```

Compared to the uninformative prior we used before, this prior assigns the highest probability to $\theta$ = 0.5. The probability gradually decreases for higher and lower values of $\theta$ (from 0.5).

```{r surfPosterior2}
posterior <- calculate_posterior(prior)
summary(posterior)
```

We can observe that the estimate of the proportion of females in the population, $\theta$, has decreased to 0.5233 (52.33%) from 0.5347 (53.47%) using the informative prior.


<br>

## Output settings

**We can set the credible level of the credible interval outputs. This is done using the `cred_level` argument.**

The default credible level is 95%.

If we want a 90% credible interval for our estimates, `cred_level=90`:

```{r surfPosterior3}
posterior <- calculate_posterior(prior, cred_level=90)
summary(posterior)
```

<br>

**We can also round the output values (in significant figures). This is done using the `signif_value` argument.**

The output below are rounded to 2 significant figures:

```{r surfPosterior4}
posterior <- calculate_posterior(prior, signif_value=2)
summary(posterior)
```

<br>

## Bivariate application

### Gender vs Qualification

**The data:**

```{r surfLikelihood2}
likelihood <- inz_lbinom(surf_data, Gender, Qualification)
```

With the two variables, we would like to estimate the proportion of females ('successes') (and hence, the proportion of males (‘failures’)) conditional on the groups in the secondary variable (Qualification).

In other words, we want to estimate the proportion of Gender for each group/level in Qualification.

**The prior:**

We will use **Beta(1,1) prior** (Uniform prior) for all groups:

```{r surfPrior3}
prior <- inz_dbeta(likelihood)
```

```{r surfPlot3, echo=FALSE}
theta <- seq(0, 1, by=0.001)

par(mfrow=c(2,2))

for (i in 1:length(likelihood$groups)) {

  y <- dbeta(theta, prior$alpha[i], prior$beta[i])

  plot(theta, y, type = "n", ylim=c(0,1.2),
       main=paste0("Prior for ", likelihood$groups[i],
                  "\nBeta(", prior$alpha[i],",", prior$beta[i],")"),
       xlab=expression(theta),
       ylab="Density", yaxs = "i")
  polygon(c(0, theta, 1),
          c(0, y, 0),
          col = "powderblue")

}
```

**The posterior:**

```{r surfPosterior5}
posterior <- calculate_posterior(prior)
summary(posterior)
```


### Changing the prior...

Let's change the prior for degree group.

We have a prior belief that the proportion of females whose highest qualification is a degree is lower than males and is somewhere around 0.35.

```{r surfPrior4}
prior <- inz_dbeta(likelihood, alpha = c(15,1,1,1), beta = c(27,1,1,1))
```

```{r surfPlot4, echo=FALSE}
theta <- seq(0, 1, by=0.001)

par(mfrow=c(2,2))

for (i in 1:length(likelihood$groups)) {

  y <- dbeta(theta, prior$alpha[i], prior$beta[i])

  plot(theta, y, type = "n", ylim=c(0, (max(y)+0.2)),
       main=paste0("Prior for ", likelihood$groups[i],
                  "\nBeta(", prior$alpha[i],",", prior$beta[i],")"),
       xlab=expression(theta),
       ylab="Density", yaxs = "i")
  polygon(c(0, theta, 1),
          c(0, y, 0),
          col = "powderblue")

}
```

```{r surfPosterior6}
posterior <- calculate_posterior(prior)
summary(posterior)
```

<br>


# Dirichlet-Multinomial

This is the parameter estimation method for when the primary variable of interest is a multi-level categorical variable. Supports single variable and two-variable (categorical/categorical) cases.

With this application we will use the 'inz_lmulti' and 'inz_ddir' functions to construct our prior and likelihood object.


## Univariate application

### Qualification variable

```{r surfQualification}
head(surf_data$Qualification, n=10)
likelihood <- inz_lmulti(surf_data, Qualification)
prior <- inz_ddir(likelihood)
```


Using a **Dirichlet(1,1,1,1) prior** (Uniform)


The posterior:
```{r surfPosterior7}
posterior <- calculate_posterior(prior)
summary(posterior)
```




<br>

### Using a differnt prior

The likelihood (data) remains the same

Let's use the Jeffreys prior - **Dirichlet(0.5,0.5,0.5,0.5)**:
```{r surfPrior5}
prior <- inz_ddir(likelihood, alpha = rep(0.5, 4))
posterior <- calculate_posterior(prior)
summary(posterior)
```

<br>

## Bivariate application

### Qualification vs Gender

```{r surfQualificationGender}
table(surf_data$Qualification, surf_data$Gender)
likelihood <- inz_lmulti(surf_data, Qualification, Gender)
prior <- inz_ddir(likelihood)
posterior <- calculate_posterior(prior)
summary(posterior)
```


### Qualification vs Ethnicity

```{r surfQualificationEthnicity}
likelihood <- inz_lmulti(surf_data, Qualification, Ethnicity)
prior <- inz_ddir(likelihood)
posterior <- calculate_posterior(prior)
summary(posterior)
```

**Changing prior**

```{r surfPrior6}
prior <- inz_ddir(likelihood,
  alpha = matrix(rep(0.5,16),
  ncol=length(likelihood$levels))
)
posterior <- calculate_posterior(prior)
summary(posterior)
```
