---
title: "Estimation Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(iNZightBayes)
```

The 'iNZightBayes' package uses simple Bayesian estimation methods to obtain summaries for some common problems, such as a sample mean or proportion, ANOVA (multiple means), and linear regression.

# Grand mean of numeric variable

Here we are calculating the mean $\mu$ and variance $\sigma^2$ of a single numeric variable $y$ with length $n$, using a Gibbs sampler. The model is
$$p(\mu,\sigma^2|y) \propto p(y|\mu,\sigma^2) p(\mu) p(\sigma^2)$$
where the prior on $\mu$ is Normal with mean $\mu_0$ and variance $\sigma_0^2$, and the prior on $\sigma^2$ is Inverse-Gamma with parameters $\alpha$ and $\beta$. The likelihood for $y$ is Normal with mean $\mu$ and variance $\sigma^2$.

Since these are conjugate, the full conditionals needed for Gibbs sampling are
$$p(\mu^{(i+1)}|\sigma^2, \ldots) = \ldots$$

# Group means of a numeric variable

# Proportions of a categorical variable

# Linear regression
